---
title: "Logistic Regression using Numerical Optimization"
author: "Erika Richter, Indrajit Nandi, Waliu Lamidi"
output: rmarkdown::Final_Project_Group9
vignette: >
  %\VignetteIndexEntry{Logistic Regression using Numerical Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteAuthor{Erika Richter, Indrajit Nandi, Waliu Lamidi}
---

## dataset: wine.csv

# Logistic Regression using Numerical Optimization

This vignette demonstrates how to use the `BinaryClassOpt` package to perform binary classification using logistic regression with numerical optimization.

## 1. Introduction

The `BinaryClassOpt` package implements logistic regression for binary classification using an optimization procedure to estimate the regression coefficients. The optimization method minimizes the negative log-likelihood function, and the package provides features like bootstrapping for confidence intervals and confusion matrix evaluation.

## 2. Installation

To install the package from GitHub, use the following command:

```{r}
devtools::install_github("emr0075/BinaryClassOpt")
```

## 3. Example Dataset
We will use a simulated dataset with binary outcomes (y) and predictors (X).

## Use wine.csv

```{r}
set.seed(42)
n <- 100  # number of samples
p <- 3    # number of predictors

# Simulating random predictors and binary response variable
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
y <- rbinom(n, 1, 0.5)  # Binary outcome
```

## 4. Running logistic regression

```{r}
# Run logistic regression
beta_hat <- logistic_regression(X, y)
beta_hat
```

## 5. Bootstrapped Confidence Intervals
```{r}
# Bootstrap confidence intervals
bootstrap_results <- bootstrap_CI(X, y, n_bootstrap = 100, alpha = 0.05)
bootstrap_results

```

## 6. Confusion Matrix and Evaluation Metrics
```{r}
# Predicted probabilities
p_hat <- sigmoid(X %*% beta_hat)

# Calculate confusion matrix and metrics
metrics <- confusion_matrix_metrics(y, p_hat, cutoff = 0.5)
metrics
```

## 7. Conclusion
```{r}
?logistic_regression
?bootstrap_CI
?confusion_matrix_metrics
```

## 8. Future Work
This package could be further enhanced by implementing additional optimization techniques, adding more metrics, or supporting multi-class classification.

## 9. References


### Step 2: Building and Previewing the Vignette

After creating this `.Rmd` file, you can build the vignette by running the following in the R console:
```{r}
devtools::build_vignettes()
```
